Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-02 Paper_Draft2/Draft2.Rmd
fixing and adding images (still having bug with one)

The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica. The raw data has 6,167 rows and each row represents an arrest charge for a defendant. AIF360’s COMPAS data includes the defendant's age, race, sex, prior charges, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest. For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American. Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two-year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot} and Table \ref{tab:recid table}) shown below.

@@

```{r age plot, echo=FALSE, fig.align='center', fig.cap= "Distribution of defendant ages by race. The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "100%"}
knitr::include_graphics("../images/age_race_plot_new.png")
```

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two-year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}
knitr::include_graphics("../images/race_recid_bar_plot_new.png")
```

>>>>

The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica. The raw data has 6,167 rows and each row represents an arrest charge for a defendant. AIF360’s COMPAS data includes the defendant's age, race, sex, prior charges, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest. For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American. Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of age (Figure \ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and two-year recidivism by race (Figure \ref{fig:recidivism mosaic}) and (Table \ref{tab:recid table}) shown below.

@@

```{r age plot, echo=FALSE, fig.align='center', fig.cap= "The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "100%"}
knitr::include_graphics("../images/age_race_plot_new.png")
```

```{r prior plot, echo=FALSE, fig.align = 'center', fig.cap = 'Black defendants, particularly men, are more likely to have a greater count of prior charges than white defendants. Male defendants have a higher number of prior charges than do female defendants. Though we do not know for sure which information goes into the COMPAS algorithm, it is likely that a defendant with prior charges will be coded as a having a higher risk of recidivism. Thus, by looking at the racial discrepancies in prior charges we can already see potential bias in the algorithm.', out.width= "100%"}
knitr::include_graphics("../images/prior_charges_plot.png")
```

```{r recidivism mosaic, echo=FALSE, fig.align = 'center', fig.cap = 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants. We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}
knitr::include_graphics("../images/recidivism_mosaic.png")
```

### 2022-05-02 Paper_Draft2/Draft2.tex
fixing and adding images (still having bug with one)

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, prior charges, what they were charged with,
and whether or not the defendant ultimately recidivated within a
two-year period after their arrest. For the purposes of our project,
which endeavors to evaluate the differing effects of the COMPAS
algorithm on white defendants versus Black defendants, we have filtered
the data to only include individuals whose race is listed as Caucasian
or African-American. Our data therefore has 5,273 rows (Figure
\ref{fig:table snip}), with the distributions of race (Figure
\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two-year
recidivism rate (Figures \ref{fig:recid plot} \&
\ref{fig:recid race plot} and Table \ref{tab:recid table}) shown below.

@@

\caption{Distribution of defendant ages by race. The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{../images/race_recid_bar_plot_new}

@@

\caption{This plot shows the distribution of two-year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
\end{figure}

\begin{longtable}[]{@{}lrrr@{}}
\caption{Incidence of recidivism by race, illustrating how a much
greater proportion (\textgreater{} 50\%) of Black defendants recidivated
than their white counterparts. \label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived & Total \\
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived & Total \\
\midrule
\endhead
African American & 1661 & 1512 & 3173 \\
Caucasian & 822 & 1278 & 2100 \\
Total & 2483 & 2790 & 5273 \\
\bottomrule
\end{longtable}

>>>>

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, prior charges, what they were charged with,
and whether or not the defendant ultimately recidivated within a
two-year period after their arrest. For the purposes of our project,
which endeavors to evaluate the differing effects of the COMPAS
algorithm on white defendants versus Black defendants, we have filtered
the data to only include individuals whose race is listed as Caucasian
or African-American. Our data therefore has 5,273 rows (Figure
\ref{fig:table snip}), with the distributions of age (Figure
\ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and
two-year recidivism by race (Figure \ref{fig:recidivism mosaic}) and
(Table \ref{tab:recid table}) shown below.

@@

\caption{The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{../images/prior_charges_plot}

@@

\caption{Black defendants, particularly men, are more likely to have a greater count of prior charges than white defendants. Male defendants have a higher number of prior charges than do female defendants. Though we do not know for sure which information goes into the COMPAS algorithm, it is likely that a defendant with prior charges will be coded as a having a higher risk of recidivism. Thus, by looking at the racial discrepancies in prior charges we can already see potential bias in the algorithm.}\label{fig:prior plot}
\end{figure}

\textbackslash{}begin\{figure\}

\{\centering \includegraphics[width=1\linewidth]{../images/recidivism_mosaic}

\}

\textbackslash{}caption\{When we divide the data into Black and white
defendants, we can see that Black defendants recidivate more than white
defendants and Black defendants are more likely to recidivate than not
recidivate. 39.14\% of white defendants did recidivate within two years
compared to 52.35\% of black defendants. We can also see that there are
more Black defendants in the dataset
overall.\}\label{fig:recidivism mosaic} \textbackslash{}end\{figure\}

\begin{longtable}[]{@{}lrrr@{}}
\caption{Incidence of recidivism by race, illustrating how a much
greater proportion (\textgreater{} 50\%) of Black defendants recidivated
than their white counterparts. \label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived &
Total\tabularnewline
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived &
Total\tabularnewline
\midrule
\endhead
African American & 1661 & 1512 & 3173\tabularnewline
Caucasian & 822 & 1278 & 2100\tabularnewline
Total & 2483 & 2790 & 5273\tabularnewline
\bottomrule
\end{longtable}
